Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-12 article.Rmd
Encode gender for AHR

```{r load-libraries, include=FALSE}
library(genderdata)
library(gender)
library(dplyr)
library(ggplot2)
library(mullenMisc)
library(knitr)
library(scales)

# TRUE to show code and warnings; FALSE to remove them for publication
display_code_blocks(FALSE)

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck’s report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period.^[http://www.historians.org/Documents/About%20AHA%20and%20Membership/CWH-Report_5.20.05.pdf]

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The “coin of the realm” for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. This process often takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the American Historical Review, the flagship journal of the American Historical Association. Along with a few dozen articles The AHR publishes five issues each year that contain, in aggregate, roughly one thousand books reviews covering (in its own words) “every major field of historical study.” The AHR is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and length of the AHR makes it an ideal source to analyze the historical profession on a large scale.

We take the AHR as a rough proxy for gatekeeping within the larger historical profession. When a book appears in the AHR it serves as a signal that other historians in positions of power have “approved” it for consideration. This signal is not necessarily correlated with the quality of the book. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession’s gatekeepers have deemed it important enough to review. On the flip side, a book that does not appear in the AHR may have gone on to have a significant impact on the profession. It is precisely this professional gatekeeping dimension that makes the AHR useful to study in the context of gender representation.

@@

[graphic]

>>>>

```{r load-libraries, include=FALSE}
library(genderdata)
library(gender)
library(dplyr)
library(ggplot2)
library(mullenMisc)
library(knitr)
library(stringr)
library(scales)

# TRUE to show code and warnings; FALSE to remove them for publication
display_code_blocks(TRUE)

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period.^[http://www.historians.org/Documents/About%20AHA%20and%20Membership/CWH-Report_5.20.05.pdf]

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. This process often takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles, the AHR publishes five issues each year that contain, in aggregate, roughly one thousand books reviews covering (in its own words) "every major field of historical study." The AHR is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and length of the AHR makes it an ideal source to analyze the historical profession on a large scale.

We take the AHR as a rough proxy for gatekeeping within the larger historical profession. When a book appears in the AHR it serves as a signal that other historians in positions of power have "approved" it for consideration. This signal is not necessarily correlated with the quality of the book. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review. On the flip side, a book that does not appear in the AHR may have gone on to have a significant impact on the profession. It is precisely this professional gatekeeping dimension that makes the AHR useful to study in the context of gender representation.

@@

```{r gender-book-authors}
ahr <- read.csv("data/ahr.utf8.csv", stringsAsFactors = FALSE) %>%
  tbl_df()

# For a number of entries the first name column contains first and middle names,
# so we're going to parse out just the first names.
ahr <- ahr %>%
  mutate(first_name = word(Author_First))

# We need to guess at the range of ages for book authors. 25 to 75 seems
# sensible enough.
min_age <- 25
max_age <- 75

# Now we will create columns for the minimum year and the maximum year for
# guessing gender.
ahr <- ahr %>%
  mutate(min_year = issueDate - max_age,
         max_year = issueDate - min_age)

# Now we need to decide which method to use when predicting the gender. In 
# general we want to use IPUMS before 1930 and SSA after 1930, but sometimes 
# we'll have to compromise. There are several kinds of obvious cases:
# 
# - If max_age < 1930, use IPUMS
# 
# - If min_age > 1930, use SSA
# 
# The tricky cases are when the terminal values overlap the ranges of one of the
# methods.
# 
# - If min_age > 1880 and max_age > 1930, use SSA
# 
# If our range of ages was greater than 50, we might have a person who we 
# guessed was born before 1880 or after 1930. But by keeping our range smaller
# than 50 (the difference between 1930 and 1880) we can avoid that problem.

# That does indeed match all cases. So now we can write a function to return the
# proper method for a given range of years.
pick_method <- function(min_age, max_age) {
  ifelse(max_age <= 1930, "ipums",
         ifelse(min_age > 1930, "ssa",
         ifelse(min_age > 1880 & max_age > 1930, "ssa", NA)))
}

# Now assign a method to the AHR dataset
ahr <- ahr %>%
  mutate(method = pick_method(min_year, max_year))

pass_through <- function(x) { return(x) }

apply_gender <- function(df) {
  gender(df$first_name, method = df$method, 
         years = c(df$min_year, df$max_year))
  }

ahr_gender <- ahr %>% 
  group_by(first_name, vol_issue_ID, Author_Full, Reviewer_First,
           Reviewer_Full, category) %>%
  do(gender_author = apply_gender(.))

# Finally we are ready to predict the gender. Because we have to combine years
# we're going to do this as a for loop, which is ugly.
# ahr_gender.l <- vector(mode = "list", nrow(ahr))

# for(i in 1:nrow(ahr)) {
#   row <- ahr[i, ]
#   result <- gender(row$first_name, method = row$method,
#                    years = c(row$min_year, row$max_year))
#   ahr_gender.l[[i]] <- result
# }

# ahr_gender.df <- ahr_gender.l %>% do.call(rbind.data.frame, .)
```
